---
title: "Covid outbreak"
author: "Johan"
date: "5/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("janitor")

library(readxl)
library(janitor)
library(tidyverse)
library(ggplot2)


read_excel_allsheets <- function(filename, tibble = FALSE) {
    # I prefer straight data.frames
    # but if you like tidyverse tibbles (the default with read_excel)
    # then just pass tibble = TRUE
    sheets <- readxl::excel_sheets(filename)
    x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X))
    if(!tibble) x <- lapply(x, as.data.frame)
    names(x) <- sheets
    x
}


```

## Excess of death and cause 


  I have only included the output of the results but you can find the R script that I made to generate this file. It can be found here. 


```{r , include=FALSE}
stmf <- readr::read_csv("https://www.mortality.org/Public/STMF/Outputs/stmf.csv", skip=1)

stmf = stmf[,1:10]


deaths <- stmf %>%
  janitor::clean_names() %>%
  gather(key = age, value = deaths,-country_code, -year, -week, -sex)%>%
  filter( sex == "b") %>% 
  mutate(
    country = recode(country_code,
      AUT = "Austria",
      BEL = "Belgium",
      DEUTNP = "Germany",
      DNK = "Denmark",
      ESP = "Spain",
      FIN = "Finland",
      GBRTENW = "England & Wales",
      ISL = "Iceland",
      NLD = "Netherlands",
      NOR = "Norway",
      PRT = "Portugal",
      SWE = "Sweden",
      USA = "United States")
  ) %>% 
  mutate(
    age = recode(age,
      d0_14 = "0-14 years",
      d15_64 = "15-64 years",
      d65_74 = "65-74 years",
      d75_84 = "75-84 years",
      d85p = "85+",
      d_total = "All")
  )  %>%
  select(year, week, country, deaths,age)
deaths <- deaths %>%
  group_by(country, year,age) %>%
  filter(!(year==2020 & week==max(week))) %>%
  select(year, week, country, age,deaths)





```


```{r adding deathstatitstics from Syria, include=FALSE}
SyriaOriginal <- read.csv("https://query.data.world/s/cpvjrlwe4oqdbm3b2ejvxtl5r4mzh6", header=TRUE, stringsAsFactors=FALSE);

SyriaFiltered = SyriaOriginal %>% filter(status == "Civilian") %>% separate( col = gender ,into = c("Age2", "Gender"), sep = " - " ) %>% select(Age2, Gender, deathdate)
SyriaFiltered$week = as.numeric(strftime(SyriaFiltered$deathdate, format = "%V"))
SyriaFiltered$year = as.numeric(strftime(SyriaFiltered$deathdate, format = "%Y"))
SyriaFiltered$country =  "Syria"


# Data  adapted from https://www.statista.com/statistics/326601/age-structure-in-syria/
# There is only information about 65+ so I distributed the to 3 percent 65-74, 1 percent 75-84 and last 0.5 percent 85+
AgeDistribution = data.frame(age  = c("0-14 years","15-64 years", "65-74 years","75-84 years","85+"), frequency= c(0.3142, 0.6408,0.03,0.01,0.005))

SyriaFilteredAdult= SyriaFiltered %>% filter(Age2 == "Adult")
adult = AgeDistribution %>% filter(age != ("0-14 years")) %>% mutate(adultFreq = frequency/sum(frequency)) 
set.seed(111)
SyriaFilteredAdult = SyriaFilteredAdult %>% 
  mutate(age = sample(x =adult$age, size = nrow(SyriaFilteredAdult), replace = T, prob = adult$adultFreq)) %>% 
  group_by(country, year, week, age) %>% 
  summarise(deaths = n())

SyriaFilteredChild = SyriaFiltered %>% filter(Age2 != "Adult") %>% mutate(age= "0-14 years") %>% 
  group_by(country, year, week, age) %>% 
  summarise(deaths = n())

Syria = rbind(SyriaFilteredChild,SyriaFilteredAdult) %>% select(year, week, country, age,deaths)
SyriaAll = Syria %>%
  group_by(year,week,country) %>%
  summarise(age = "All", deaths = sum(deaths) )

SyriaDeaths = rbind(Syria,SyriaAll)



```


```{r adding FHM Covid 19 report for sweden, include=FALSE}




destfile = "Folkhalsomyndigheten_Covid19_2.xlsx"
curl::curl_download("https://www.arcgis.com/sharing/rest/content/items/b5e7488e117749c19881cce45db13f7e/data", destfile = destfile)
FHMdata = read_excel_allsheets(filename = destfile)


#Get reported cases per day
CasesPerDay = FHMdata[[1]]
CasesPerDay = CasesPerDay[(1:nrow(CasesPerDay)-1), ]


#Get reported deaths per day
CovidDeaths= FHMdata[[2]]
CovidDeaths = CovidDeaths[(1:nrow(CovidDeaths)-1), ]

# For some reason that I dont understand the dates get wrong in the puling from excel and not in the other case. Not spending time on this but just copying for one to the other. 
CovidDeaths$Datum_avliden = CasesPerDay$Statistikdatum[(nrow(CasesPerDay)-nrow(CovidDeaths)+1):nrow(CasesPerDay)]
CovidDeaths$week = as.numeric(strftime(CovidDeaths$Datum_avliden, format = "%V"))
CovidDeaths$year = 2020

CovidDeathsPerWeek = CovidDeaths %>% group_by(year,week) %>% summarise(deaths = sum(Antal_avlidna))


# removing last week since it is uncertain
CovidDeathsPerWeek = CovidDeathsPerWeek[(1:nrow(CovidDeathsPerWeek)-1), ]
CovidDeathsPerWeek$excess = CovidDeathsPerWeek$deaths

CovidDeathsPerWeek$country = "Sweden"
CovidDeathsPerWeek$median_deaths = 0
CovidDeathsPerWeek$age = "All"

CovidDeathsPerWeek = CovidDeathsPerWeek %>% select(year,week,country, age, deaths, median_deaths,excess )



```


```{r merging the data, include=FALSE }





recent_deaths <- deaths %>%
  filter(year >= 2011 & year <= 2019) %>%
  group_by(country,week,age) %>%
  summarise(median_deaths = median(deaths)) %>%
  ungroup()
excess_deaths <- deaths %>%
  filter(year >= 2011) %>%
  left_join(recent_deaths) %>%
  mutate(excess = deaths - median_deaths)



war_deaths <- Syria %>%
  filter(year >= 2011 & year <= 2019) %>%
  group_by(country,week,age) %>%
  summarise(median_deaths = sum(deaths)) %>%
  ungroup()

excess_deaths_Syria <- Syria %>%
   filter(year >= 2011) %>%
  left_join(war_deaths) %>%
  mutate(excess = deaths )



SummaryWarSyriah = war_deaths %>% mutate(year = 2020,deaths = median_deaths,excess = median_deaths) %>% select(year, week, country, age,deaths, median_deaths,excess)

SummaryWarSyriah$reason = "War"
excess_deaths_Syria$reason = "War"
CovidDeathsPerWeek$reason = "Covid-19"
excess_deaths$reason = "Everything"


```



```{r merge all the datasets to compare them, include=FALSE }

excess_deaths = rbind(excess_deaths,excess_deaths_Syria)
excess_deaths = rbind(data.frame(excess_deaths),SummaryWarSyriah)
excess_deaths = rbind(data.frame(excess_deaths),data.frame(CovidDeathsPerWeek))
excess_deaths = excess_deaths %>% mutate(ageShort = recode(age,
                                                               "65-74 years" = "65+ years",
                                                               "75-84 years" = "65+ years",
                                                               "85+" = "65+ years")) 
excess_deaths2  =  excess_deaths %>% 
  group_by(year,week,country,ageShort,reason) %>%
  summarize(excess = sum(excess), nr = n()) %>% rename("age" = ageShort)



unique(excess_deaths2$country)

```





```{r visualize the data, include=FALSE}
excess_deaths_Final  = excess_deaths2%>% filter( country != "United States" ) 
thisYearData = excess_deaths_Final %>% filter( year == 2020 ) 

ExcessDeathPlot = excess_deaths_Final %>% filter(age == "All")%>% filter(reason == "Everything") %>%
  mutate(thisyear = (year == 2020)) %>%
  ggplot(aes(x=week, y=excess, group=year)) +
  geom_hline(yintercept=0, col='gray') +
  geom_line(aes(col=thisyear, linetype = reason)) +
  facet_wrap(~ country, scales='free_y') +
  scale_color_manual(values=c("FALSE"='gray',"TRUE"='red', "d0_14" = 'yellow',"d15_64" = 'blue',"d65_74"='orange',"d75_84"='green',"d85p" = "black")) +
  ggtitle("Weekly excess deaths")


SeperateByAge = thisYearData %>% filter( week < 52)%>%
  ggplot(aes(x=week, y=excess,color = reason)) +
  geom_line() +
  facet_grid(age ~country, scales='free_y') +
  ggtitle("Weekly excess deaths")


SeperateByAgeNordic = thisYearData %>% filter( week < 52) %>%
  filter( country %in% c("Sweden","Norway","Denmark", "Syria")) %>%
  ggplot(aes(x=week, y=excess,color = reason)) +
  geom_line() + 
  facet_grid(age ~country) +
  ggtitle("Weekly excess deaths in 2020 compared to average of past 8 years")




countryPlots = list()
for (coun in unique(excess_deaths_Final$country)){
  countryPlots[[coun]] = excess_deaths_Final %>% filter(country == coun & week < 25)%>%
  mutate(thisyear = (year == 2020)) %>%
  ggplot(aes(x=week, y=excess, group=year)) +
  geom_hline(yintercept=0, col='gray') +
  geom_line(aes(col=thisyear)) +
  facet_wrap(~ age, scales='free_y') +
  scale_color_manual(values=c("FALSE"='gray',"TRUE"='red')) +
  guides(col=FALSE) +
  ggtitle(paste("Number of excess death per week in", coun, "\nto average excess deathrate for the last eight years")
)
}

agePlots = list()
for (AGE in unique(excess_deaths_Final$age)){
  agePlots[[AGE]] = excess_deaths_Final %>% filter(age == AGE & week < 25)%>%
  mutate(thisyear = (year == 2020)) %>%
  ggplot(aes(x=week, y=excess, group=year)) +
  geom_hline(yintercept=0, col='gray') +
  geom_line(aes(col=thisyear)) +
  facet_wrap(~ country, scales='free_y') +
  scale_color_manual(values=c("FALSE"='gray',"TRUE"='red')) +
  guides(col=FALSE) +
  ggtitle(paste("Number of excess death per week in age group", AGE, "\nto average excess deathrate for the last eight years")
)
}
```



```{r Plot excess death over time }

ExcessDeathPlot




```

### Death seperated by age, cause and country in Europe 

This is given in one plot. It is quite big. I recomend looking at the pdf available here() 
```{r This year divided into age groups and cause of death Europe} 

SeperateByAge
ggsave("../figures/ExcessDeath_Europe.pdf", width = 15, height = 10)

```


### Death seperated by age, cause and country in Nordic countries 


This is the same plot with only the nordic countries and relating to the sum of deats caused in Syria over the last 9 years. I think it gives some perspective. The red line is based on the reported deaths from the FHM. The blue line is based on a database maintained by the Violations Documentation Center in Syria. Data can be found here 
( https://data.world/polymathic/casualties-of-the-syrian-civil-war/workspace/file?filename=VDC_Syria_CASREP.csv)



```{r This year divided into age groups and cause of death Nordic}

SeperateByAgeNordic 

```

## Plot excess of death in different countries


```{r countryPlots}

countryPlots
```

## Plot excess of death in different age categories

```{r agePlots}

agePlots
```


